Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory

### 2022-12-04 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

\caption{Mutation bias and GC-content shape protein composition. Panels \textbf{(A)}, \textbf{(B)} and \textbf{(C)} show expected frequency, gain probability (\autoref{aagaineq}), and loss probability (\autoref{aalosseq}) for different amino acids (horizontal axes). The vertical axis in all these three subplots show the normalized values of these measures such that their sum over all amino acids is equal to one. Panel \textbf{(D)} shows the directionality (colored tiles, $\textit{D}_{x\to y}$, \autoref{directionality}) of the substitution from an amino acid (vertical axis) to another (horizontal axis). Positive values of $\textit{D}_{x\to y}$ are shown in shades of red, whereas negative values are shown in shades of blue.}
\label{aafreq}

@@

A previous study found that mutations that cause hydrophobic amino acids to appear on protein surfaces, can lead to evolution of protein dimers \citep{hydrophobicRatchet}. The authors \citep{hydrophobicRatchet} also suggested that such an evolutionary process may be widespread because mutation bias tends to facilitate emergence of hydrophobic amino acids. They based this argument on another study that estimated mutation rate bias in bacteria \citep{bacteriaMutbias}. The authors also argued that hydrophobic amino acids are more frequently found in random protein sequences.

@@

We found that the average change in hydrophobicity to be $-0.033$, which suggests that on an average mutations cause a very small increase in hydrophobicity. If we exclude amino acid pairs where both amino acids are either hydrophobic or are hydrophilic, then the average change in hydrophobicity is even smaller ($-3.7\times 10^{-4}$).

@@

In this work, we addressed some fundamental questions about \textit{de novo} gene evolution. Broadly, we asked how quickly these genes can emerge and be lost, whether their birth and death has a preferred trajectory, and how the composition of their protein sequence is determined. To this end, we developed mathematical models and used them to address specific questions. These models are based on mutation probability and mutation rate bias, and represent neutral evolution where DNA sequences are not evolving under natural selection. Thus our models provide an opportunity to test the alternative hypothesis that some \textit{de novo} genes are evolving under selection.

@@

We answer the long standing question of what is the sequence of \textit{de novo} emergence. That is, whether ORF emerges first or transcription emerges first \citep{EBB-F1000}. We show that, in the absence of selection, long ORFs are likely to emerge in existing transcripts (transcription emerges first), whereas short ORFs are likely to emerge before transcription emerges in the genomic locus. Long non-coding RNAs may be a potential source of \textit{de novo} genes with long ORFs but this evolutionary process may be constrained by a different set of factors. A previous study has shown that new transcripts frequently arise in regions overlapping existing genes. These new transcripts can harbor ORFs which in turn leads to \textit{de novo} gene emergence \citep{albaYeastdenovo}. This is an example of RNA-first trajectory. However, the ORFs present in these genes may be under indirect selection pressure that acts on the overlapping gene sequences. Therefore, they may be less frequently mutated or lost.

>>>>

\caption{Mutation bias and GC-content shape protein composition. Panels \textbf{(A)}, \textbf{(B)} and \textbf{(C)} show expected frequency, gain probability (\autoref{aagaineq}), and loss probability (\autoref{aalosseq}) for different amino acids (horizontal axes). The vertical axis in all these three subplots show the normalized values of these measures such that their sum over all amino acids is equal to one. Panel \textbf{(D)} shows the directionality (colored tiles, $\textit{D}_{x\to y}$, \autoref{directionality}) of the substitution from an amino acid (vertical axis) to another (horizontal axis). Positive values of $\textit{D}_{x\to y}$ are shown in shades of red, whereas negative values are shown in shades of blue. In all panels, amino acids are ordered based on their chemical similarity \citep{PMBEC}.}
\label{aafreq}

@@

A previous study found that mutations that cause hydrophobic amino acids to appear on protein surfaces, can lead to evolution of protein dimers \citep{hydrophobicRatchet}. The authors also suggested that such an evolutionary process may be widespread because mutation bias tends to facilitate emergence of hydrophobic amino acids. They based this argument on another study that estimated mutation rate bias in bacteria \citep{bacteriaMutbias}. The authors \citep{hydrophobicRatchet} further suggested that hydrophobic amino acids are more frequently found in random protein sequences.

@@

We found the average change in hydrophobicity to be $-0.033$, which suggests that on an average mutations cause a very small increase in hydrophobicity. We reiterate that hydrophobic amino acids have a negative value hydrophobicity and thus a negative change in hydrophobicity also denotes a shift towards a more hydrophobic protein. To focus exclusively on non-hydrophobic to hydrophobic amino acid substitutions, we excluded amino acid pairs where both amino acids are either hydrophobic or are non-hydrophobic. That is, we only analysed pairs of amino acids that have opposite signs of hydrophobicity. With this focused analysis, we find that the average change in hydrophobicity is even smaller ($-3.7\times 10^{-4}$).

@@

In this work, we addressed some fundamental questions about \textit{de novo} emergence of protogenes. Broadly, we asked how quickly these genes can emerge and be lost, whether their birth and death has a preferred trajectory of mutational events, and how the composition of their protein sequence is determined. To this end, we developed mathematical models and used them to address specific questions. These models are based on mutation probability and mutation rate bias, and represent neutral evolution where DNA sequences are not evolving under natural selection. Thus our models provide an opportunity to test the alternative hypothesis that some \textit{de novo} genes are evolving under selection.

@@

Witin the limitations of our model assumptions, we answer the long standing question of which trajectory of mutational events leads to \textit{de novo} emergence. That is, whether ORF emerges first or transcription emerges first \citep{EBB-F1000}. We show that, in the absence of selection, long ORFs are likely to emerge in existing transcripts (transcription emerges first), whereas short ORFs are likely to emerge before transcription emerges in the genomic locus. Long non-coding RNAs may be a potential source of \textit{de novo} genes with long ORFs but this evolutionary process may be constrained by a different set of factors. A previous study has shown that new transcripts frequently arise in regions overlapping existing genes. These new transcripts can harbor ORFs which in turn leads to \textit{de novo} gene emergence \citep{albaYeastdenovo}. This is an example of RNA-first trajectory. However, the ORFs present in these genes may be under indirect selection pressure that acts on the overlapping gene sequences. Therefore, they may be less frequently mutated or lost.
